Rising from the Ashes II: The Bar-driven Abundance Bimodality of the Milky Way

### 2024-10-11 paper/ms.tex
Fix abstract
Lars edits

\title{Rising from the Ashes II: The Bar-driven Abundance Bimodality in the Milky Way}

@@

\begin{abstract}
    The Milky Way hosts at least two modes in its present day distribution of Fe and $\alpha$-elements. The exact cause of this bimodality is disputed, but one class of explanations involves the merger between the Milky Way and a relatively massive satellite (Gaia-Sausage-Enceladus) at $z\sim2$. However, reproducing this bimodality in simulations is not straightforward, with conflicting results on the prevalance, morphology, and mechanism behind multimodality. We present a case study of a galaxy in the Illustris TNG50 simulation which undergoes a sequence of starburst, brief quiescence, and then rejuvenation.

    Following a minor post-processing adjustment -— motivated by recent findings on the star formation efficiency in TNG -- which increases the \alphaFe{} values of older star particles, we demonstrate that this galaxy exhibits a pronounced bimodal distribution in the \alphaFe{}-\FeH{} plane.

    This sequence results in a pronounced abundance bimodality after a post-processing adjustment of the \alphaFe{} of old stars which is designed to mimic a higher star formation efficiency in dense gas.

    The high- and low-$\alpha$ sequences are neatly separated in time by the brief quiescent period. The quiescent period in this galaxy is not associated with a merger but by the formation of a bar followed by AGN activity. This galaxy indicates a novel scenario in which the $\alpha$-bimodality in the Milky Way is caused by AGN-induced quenching triggered by the formation of its bar. We predict that abundance bimodalities should be more common in barred as opposed to unbarred galaxies.
  \end{abstract}

@@

In \citet{2024arXiv240707985B} (hereafter Paper~I), we proposed an alternate scenario for the formation of the bimodality, driven by a brief ($\sim300\,\Myr$) quiescent period in the Galaxy's history (an observational claim with a broader period of quiescence was made in \citet{2016A&A...589A..66H}). This period has two effects. First, the reduced SFR lowers the number of $\alpha$-producing Type~II SNe relative to Type~Ia\footnote{These SNe occur on different timescales (10s of Myr after star formation for Type~II, as compared to 100s of Myr to Gyrs for Type~Ia), so if the SFR drops, the number of Type~II SNe drops faster than Type~Ia.}, causing a sharp decline in the \alphaFe{} ratio in the star-forming gas. Second, the lower SFR results in fewer stars forming in the region between the high- and low-$\alpha$ sequences, reducing the occurance of this transitional population in present-day observations. The combination of these two effects creates the valley between the high- and low-$\alpha$ sequences. This mechanism resembles two-phase infall models that incorporate a temporary halt in star formation \citep[][and references therein]{2024arXiv240511025S}, though the quiescent period in Paper~I is much shorter.

@@

\section{Methods}\label{sec:methods}
\subsection{IllustrisTNG Sample}\label{ssec:tng}
We use Illustris TNG50 \citep{2019MNRAS.490.3196P, 2019MNRAS.490.3234N}, a cosmological simulation of a $\sim50\,\textrm{cMpc}$ box at high resolution ($m_{\textrm{baryon}}\sim8.5\times10^4\,\Msun$). It uses the gravito-magneto-hydrodynamics code \texttt{AREPO} \citep{2010MNRAS.401..791S, 2016MNRAS.455.1134P}, along with the TNG model \citep{2013MNRAS.436.3031V, 2017MNRAS.465.3291W, 2018MNRAS.473.4077P}. This model includes several subgrid processes, including a wind generation model, chemical enrichment from SNe and asymptotic giant branch stars, and thermal and kinetic feedback from AGN.

@@

% Please add the following required packages to your document preamble:
\begin{table}[]
  \centering
  \begin{tabular}{@{}ll@{}}
  \toprule
  \multicolumn{2}{c}{dark matter}                                                  \\ \midrule
  \multicolumn{1}{l|}{$M_{200}$}                      & $4.97\times10^{12}\,\Msun$ \\
  \multicolumn{1}{l|}{$R_{200}$}                      & $532\,\kpc$                \\
                                                      &                            \\
  \multicolumn{2}{c}{baryons}                                                      \\ \midrule
  \multicolumn{1}{l|}{$\rhalf$}                       & $2\,\kpc$                  \\
  \multicolumn{1}{l|}{$M_{\textrm{star}}(r<4\rhalf)$} & $7.6\times10^{12}\,\Msun$  \\
  \multicolumn{1}{l|}{$M_{\textrm{gas}}(r<4\rhalf)$}  & $2.6\times10^{11}\,\Msun$  \\
  \multicolumn{1}{l|}{$\textrm{SFR}$}                 & $0.56\,\Msunyr$            \\
  \multicolumn{1}{l|}{$M_{\textrm{BH}}$}              & $2.9\times10^{8}\,\Msun$   \\
  $\dot{M}_{\textrm{BH}}/\dot{M}_{\textrm{Edd}}$      & $4.5\times10^{-5}$         \\ \bottomrule
  \end{tabular}
  \caption{Summary statistics of our subhalo of interest at $z=0$. $M_{200}$ and $R_{200}$ are relative to the mean density of the universe, the SFR is for all gas bound to the subhalo, and the BH refers to the central BH.}
  \label{tab:summ}
  \end{table}

@@

In the SoI, there is an age gap at $sim10.6\,\Gyr$, which we mark with a vertical dashed line in the upper row. Star particles older than this line have a much clearer gradient in time with \MgFe{} than stars that form after. In the $\FeH=-0.25$ bin, star particles which form directly after this line have a slightly reduced \MgFe{} than stars which form a short time later.

@@

In Figure~\ref{fig:fig1}, we showed that a time-linear $\alpha$-enhancement of older stars (formed before $z=1.5$) leads to the emergence of a pronounced chemical bimodality. This $\alpha$-enhancement corresponds to a more rapid decline in \alphaFe{} over time at high redshifts. 
We demonstrate the physical motivation behind this $\alpha$-enhnacement in Figure~\ref{fig:vice}.

@@

\subsection{Cause of Bimodality}\label{ssec:bim_cause}
Here we argue that the available evidence for the cause of the bimodality is consistent with the scenario presented in Paper~I. They presented an idealized simulation resembling the merger between the Milky Way and GSE. They varied the orbital parameters slightly in a grid of 27 simulations and found that simulations which had a brief quiescent period as a result of the merger led to a bimodal abundance distribution. They argued that AGN activity was responsible for this brief quenching.

Once the $\alpha$-enhancement post-processing has been done, the SoI that we have studied in this work is consistent with this scenario. The vertical dashed line in Figures~\ref{fig:alpha} and \ref{fig:history} denotes the transition between the high- and low-$\alpha$ sequences. We place it at a lookback time of $10.6\,\Gyr$ or a cosmic time of $\sim3.2\,\Gyr$ or $z\sim2.05$.\footnote{The transition between the high- and low-$\alpha$ sequences occurs approximately $1\,\Gyr$ before $z=1.5$, where the $\alpha$-enhancement begins. This, combined with the fact that not all of the subhalos in our sample display bimodalities (Appendix~\ref{app:rand_fig1}), shows that the $\alpha$-enhancement is not the direct cause of the bimodality.} The upper right panel of Figure~\ref{fig:alpha} shows 2D histograms of \MgFe{} vs star particle age in fixed bins of \FeH{} (each color is a different \FeH{} bin, and an offset is given to \MgFe{} to prevent overlap). Stars formed before the dashed line had a steep decline of \MgFe{} with time, while the relation in star particles formed after the dashed line is much more flat. At the $\alpha$-transition, we see a gap in the $\FeH=-0.5$ and $-0.25$ bins.

@@

The \MgFe{} value of star forming gas is the result of a complicated mixture of many different aspects of the TNG model, to name a few: stellar and AGN feedback which alter gas inflows and outflows, secular, dynamical evolution, SF prescription, magnetic fields, (lack of) cosmic rays, diffusivity of hydrodynamics solver, and, of course, enrichment models. Isolating the cause of the ``incorrect''\footnote{Our case study of a single subhalo, selected in a non-reproducible manner, is hardly cause to firmly assert that the fiducial evolution in TNG50 is incorrect.} \MgFe{} vs time evolution at high-$z$ is not straightforward nor, in our opinion, even possible. However, we do offer one reasonable explanation: the SFE at high densities, more present at high-$z$, is too low.

We demonstrate the impact of the SFE on the \alphaFe{} ratio using a simple one-zone chemical evolution model with the publicly available code \texttt{VICE}. The details of our setup is given in Section~\ref{ssec:onezone_met}. We vary the SFE of the model ($\dot{Sigma}_{\textrm{star}}/\Sigma_{\textrm{gas}}$), and examine the impact on the \MgFe{} ratio as a function of time. We find that a higher SFE does lead to a more rapid reduction in \MgFe{}. The rate of decrease in \MgFe{}, at its maximum, varies between $\sim-0.25$, $-0.35$, and $-0.5\,\dex/\Gyr$ in the $\textrm{SFE}=0.2$, $1$, and $2\,\Gyr^{-1}$ models, respectively. For our post-processing, we assumed an additional decrease rate of $0.1\,\dex/\Gyr$. Such a difference is well within the range of \MgFe{} slopes seen in our different $\tau_{\star}$ models, implying a factor of only $\sim2$ to $5$ in the SFE is needed to reach our post-processing slope.

@@

An intuitive understanding of the impact the decline in \alphaFe{} vs time has is that, when \alphaFe{} declines rapidly, it is a better estimator of age. When \alphaFe{} is a better estimator of age, events which are separated temporally become better separated in the abundance plane.

\subsection{Direct Comparison to Observations}\label{ssec:compare_obs}
The lower left panel of Figure~\ref{fig:alpha} shows the \MgFe{} vs age of stars in bins of \FeH{} which pass our solar neighborhood selection and are present in the APOKASC-3 catalog. In the bins where the bimodality is strongest (blue and orange, $\FeH=-0.5$ and $-0.25$, respectively), we see that there is a sort of two tiered distribution with significant overlap in age. In the SoI with age and abundance errors shown in the lower right panel, a similar distribution can be seen in all \FeH{} bins. An examination of the true distribution in the SoI (upper right panel), we see that the two distributions are in fact very cleanly separated in age.

@@

This bimodality can be traced to an event that occurred at an age of $10.6\,\Gyr$, or formation time of $\sim3.2\,\Gyr$ ($z\sim2$). Shown as a vertical dashed line in Figure~\ref{fig:alpha}, star particles that form before this time are at high-\MgFe{}, while star particles that form after are at low-\MgFe{}. This time is associated with a global reduction in the SFR (Figure~\ref{fig:history}), which appears to be related to AGN activity induced by the formation of a bar (Figure~\ref{fig:seq}).

We argue that the reduction in the SFR has two effects which lead to the $\alpha$-bimodality. First, a lower production of Mg relative to Fe is the natural result of a lack of Type~II relative to Type~Ia SNe. The latter remain high due to their longer timescales, expediting a drop in \MgFe{}. Second, the lack of SF means that stars at \MgFe{} intermediate between the high- and low-$\alpha$ sequences never form. This argument is the same as presented in Paper~I and is consistent with the one-zone chemical evolution models of \citet{2020MNRAS.498.1364J}. When observational errors are added to the subhalo's abundance vs age distribution, we show that it is broadly consistent with the distribution observed in the Milky Way (Figure~\ref{fig:alpha} and Section~\ref{ssec:compare_obs}).

>>>>

\title{Rising from the Ashes II: The Bar-driven Abundance Bimodality of the Milky Way}

@@

\begin{abstract}
    The Milky Way hosts at least two modes in its present day distribution of Fe and $\alpha$-elements. The exact cause of this bimodality is disputed, but one class of explanations involves the merger between the Milky Way and a relatively massive satellite (Gaia-Sausage-Enceladus) at $z\sim2$. However, reproducing this bimodality in simulations is not straightforward, with conflicting results on the prevalance, morphology, and mechanism behind multimodality. We present a case study of a galaxy in the Illustris TNG50 simulation which undergoes a sequence of starburst, brief quiescence, and then rejuvenation.
    This sequence results in a pronounced abundance bimodality after a post-processing adjustment of the \alphaFe{} of old stars which is designed to mimic a higher star formation efficiency in dense gas.
    The high- and low-$\alpha$ sequences are neatly separated in time by the brief quiescent period. The quiescent period in this galaxy is not associated with a merger but by the formation of a bar followed by AGN activity. This galaxy indicates a novel scenario in which the $\alpha$-bimodality in the Milky Way is caused by AGN-induced quenching triggered by the formation of its bar. We predict that abundance bimodalities should be more common in barred as opposed to unbarred galaxies.
  \end{abstract}

@@

In \citet{2024arXiv240707985B} (hereafter Paper~I), we proposed an alternate scenario for the formation of the bimodality, driven by a brief ($\sim300\,\Myr$) quiescent period in the Galaxy's history (an observational claim with a broader period of quiescence was made in \citet{2016A&A...589A..66H}). This period has two effects. First, the reduced SFR lowers the number of $\alpha$-producing Type~II SNe relative to Type~Ia\footnote{These SNe occur on different timescales (10s of Myr after star formation for Type~II, as compared to 100s of Myr to Gyrs for Type~Ia), so if the SFR drops, the number of Type~II SNe drops faster than Type~Ia.}, causing a sharp decline in the \alphaFe{} ratio in the star-forming gas. Second, the lower SFR results in fewer stars forming in the region between the high- and low-$\alpha$ sequences, reducing the occurrence of this transitional population in present-day observations. The combination of these two effects creates the valley between the high- and low-$\alpha$ sequences. This mechanism resembles two-phase infall models that incorporate a temporary halt in star formation \citep[][and references therein]{2024arXiv240511025S}, though the quiescent period in Paper~I is much shorter.

@@

\section{Methods}\label{sec:methods}
\subsection{IllustrisTNG Sample}\label{ssec:tng}
We use Illustris TNG50 \citep{2019MNRAS.490.3196P, 2019MNRAS.490.3234N, 2019ComAC...6....2N}, a cosmological simulation of a $\sim50\,\textrm{cMpc}$ box at high resolution ($m_{\textrm{baryon}}\sim8.5\times10^4\,\Msun$). It uses the gravito-magneto-hydrodynamics code \texttt{AREPO} \citep{2010MNRAS.401..791S, 2016MNRAS.455.1134P}, along with the TNG model \citep{2013MNRAS.436.3031V, 2017MNRAS.465.3291W, 2018MNRAS.473.4077P}. This model includes several subgrid processes, including a wind generation model, chemical enrichment from SNe and asymptotic giant branch stars, and thermal and kinetic feedback from AGN.

@@

% Please add the following required packages to your document preamble:
\begin{deluxetable}{lc}
  \tablecaption{Summary statistics of our subhalo of interest at $z=0$. $M_{200}$ and $R_{200}$ are relative to the mean density of the universe, the SFR is for all gas bound to the subhalo, and the BH refers to the central BH. \label{tab:summ}}
  \tablewidth{0pt}
  \tablehead{
    % \colhead{} & \colhead{}
  }
  \startdata
  \multicolumn{2}{c}{\textbf{dark matter}} \\ \hline
  $M_{200}$ & $4.97\times10^{12}\,\Msun$ \\
  $R_{200}$ & $532\,\kpc$ \\
  & \\
  \multicolumn{2}{c}{\textbf{baryons}} \\ \hline
  $\rhalf$ & $2\,\kpc$ \\
  $M_{\textrm{star}}(r<4\rhalf)$ & $7.6\times10^{12}\,\Msun$ \\
  $M_{\textrm{gas}}(r<4\rhalf)$ & $2.6\times10^{11}\,\Msun$ \\
  $\textrm{SFR}$ & $0.56\,\Msunyr$ \\
  $M_{\textrm{BH}}$ & $2.9\times10^{8}\,\Msun$ \\
  $\dot{M}_{\textrm{BH}}/\dot{M}_{\textrm{Edd}}$ & $4.5\times10^{-5}$ \\
  \enddata
\end{deluxetable}

@@

In the SoI, there is an age gap at $\sim10.6\,\Gyr$, which we mark with a vertical dashed line in the upper row. Star particles older than this line have a much clearer gradient in time with \MgFe{} than stars that form after. In the $\FeH=-0.25$ bin, star particles which form directly after this line have a slightly reduced \MgFe{} than stars which form a short time later.

@@

In Figure~\ref{fig:fig1}, we showed that a time-linear $\alpha$-enhancement of older stars (formed before $z=1.5$) leads to the emergence of a pronounced chemical bimodality. This $\alpha$-enhancement corresponds to a more rapid decline in \alphaFe{} over time at high redshifts. 
We demonstrate the physical motivation behind this $\alpha$-enhancement in Figure~\ref{fig:vice}.

@@

\subsection{Cause of Bimodality}\label{ssec:bim_cause}
We argue that the available evidence for the cause of the bimodality aligns with the scenario outlined in Paper~I. That study presented an idealized simulation resembling the merger between the Milky Way and GSE, where the orbital parameters were varied slightly across a grid of 27 simulations. The results showed that simulations featuring a brief quiescent period due to the merger produced a bimodal abundance distribution. In this scenario, gas expulsion by AGN activity was identified as the driving force behind the quenching.

Once the $\alpha$-enhancement post-processing has been done, the SoI that we have studied in this work is consistent with this scenario. The vertical dashed line in Figures~\ref{fig:alpha} and \ref{fig:history} denotes the transition between the high- and low-$\alpha$ sequences. We place it at a lookback time of $10.6\,\Gyr$ or a cosmic time of $\sim3.2\,\Gyr$ or $z\sim2.05$.\footnote{The transition between the high- and low-$\alpha$ sequences occurs approximately $1\,\Gyr$ before $z=1.5$, where the $\alpha$-enhancement begins. This, combined with the fact that not all of the subhalos in our sample display bimodalities (Appendix~\ref{app:rand_fig1}), shows that the $\alpha$-enhancement is not the direct cause of the bimodality.} The upper right panel of Figure~\ref{fig:alpha} shows 2D histograms of \MgFe{} vs. star particle age in fixed bins of \FeH{} (each color is a different \FeH{} bin, and an offset is given to \MgFe{} to prevent overlap). Stars formed before the dashed line had a steep decline of \MgFe{} with time, while the relation in star particles formed after the dashed line is much more flat. At the $\alpha$-transition, we see a gap in the $\FeH=-0.5$ and $-0.25$ bins.

@@

The \MgFe{} value of star forming gas is the result of a complicated mixture of many different aspects of the TNG model, to name a few: stellar and AGN feedback which alter gas inflows and outflows, secular, dynamical evolution, SF prescription, magnetic fields, (lack of) cosmic rays, diffusivity of hydrodynamics solver, and, of course, enrichment models. Isolating the cause of the ``incorrect''\footnote{Our case study of a single subhalo, selected in a non-reproducible manner, is hardly cause to firmly assert that the fiducial evolution in TNG50 is incorrect.} \MgFe{} vs. time evolution at high-$z$ is not straightforward nor, in our opinion, even possible. However, we do offer one reasonable explanation: the SFE at high densities, more present at high-$z$, is too low.

We demonstrate the impact of the SFE on the \alphaFe{} ratio using a simple one-zone chemical evolution model with the publicly available code \texttt{VICE}. The details of our setup is given in Section~\ref{ssec:onezone_met}. We vary the SFE of the model ($\dot{\Sigma}_{\textrm{star}}/\Sigma_{\textrm{gas}}$), and examine the impact on the \MgFe{} ratio as a function of time. We find that a higher SFE does lead to a more rapid reduction in \MgFe{}. The rate of decrease in \MgFe{}, at its maximum, varies between $\sim-0.25$, $-0.35$, and $-0.5\,\dex/\Gyr$ in the $\textrm{SFE}=0.2$, $1$, and $2\,\Gyr^{-1}$ models, respectively. For our post-processing, we assumed an additional decrease rate of $0.1\,\dex/\Gyr$. Such a difference is well within the range of \MgFe{} slopes seen in our different $\tau_{\star}$ models, implying a factor of only $\sim2$ to $5$ in the SFE is needed to reach our post-processing slope.

@@

An intuitive understanding of the impact the decline in \alphaFe{} vs. time has is that, when \alphaFe{} declines rapidly, it is a better estimator of age. When \alphaFe{} is a better estimator of age, events which are separated temporally become better separated in the abundance plane.

\subsection{Direct Comparison to Observations}\label{ssec:compare_obs}
The lower left panel of Figure~\ref{fig:alpha} shows the \MgFe{} vs. age of stars in bins of \FeH{} which pass our solar neighborhood selection and are present in the APOKASC-3 catalog. In the bins where the bimodality is strongest (blue and orange, $\FeH=-0.5$ and $-0.25$, respectively), we see that there is a sort of two tiered distribution with significant overlap in age. In the SoI with age and abundance errors shown in the lower right panel, a similar distribution can be seen in all \FeH{} bins. An examination of the true distribution in the SoI (upper right panel), we see that the two distributions are in fact very cleanly separated in age.

@@

This bimodality can be traced to an event that occurred at an age of $10.6\,\Gyr$, or cosmic time of $\sim3.2\,\Gyr$ ($z\sim2$). Shown as a vertical dashed line in Figure~\ref{fig:alpha}, star particles that form before this time are at high-\MgFe{}, while star particles that form after are at low-\MgFe{}. This time is associated with a global reduction in the SFR (Figure~\ref{fig:history}), which appears to be related to AGN activity induced by the formation of a bar (Figure~\ref{fig:seq}).

We argue that the reduction in the SFR has two effects which lead to the $\alpha$-bimodality. First, a lower production of Mg relative to Fe is the natural result of a lack of Type~II relative to Type~Ia SNe. The latter remain high due to their longer timescales, expediting a drop in \MgFe{}. Second, the lack of SF means that stars at \MgFe{} intermediate between the high- and low-$\alpha$ sequences never form. This argument is the same as presented in Paper~I and is consistent with the one-zone chemical evolution models of \citet{2020MNRAS.498.1364J}. When observational errors are added to the subhalo's abundance vs. age distribution, we show that it is broadly consistent with the distribution observed in the Milky Way (Figure~\ref{fig:alpha} and Section~\ref{ssec:compare_obs}).
